The Gamma: Data Exploration through Iterative Prompting

### 2021-03-23 wip-design/paper.tex
Simplifying case study discussion

\begin{abstract}
  Data exploration tools based on code have many desirable characteristics. They can easily access
  a wide range of different data sources, result in reproducible scripts and encourage users to
  reuse and modify existing data analyses. Unfortunately, most programming tools require expert
  coding skills. Can we make data exploration based on code accessible to non-experts?
  %
  We present The Gamma, a novel programmatic data exploration environment.
  The Gamma is based on a single interaction principle and using it results in transparent and
  reproducible scripts. This lowers the barrier to entry and allows learning from previously
  created data analyses. We evaluate the usability and learnability of The Gamma through a user
  study on non-technical employees of a research institute.
  %
  Our work shows that we may not need to shy away from code in order to build accessible,
  reproducible and transparent tools that will allow a broad audience to benefit from the
  rise of open data.
\end{abstract}

@@

We contribute The Gamma, a programmatic data exploration environment for non-experts. The Gamma
is based on a single, easy to understand interaction principle and provides a uniform
access to a range of data sources including data tables, graph databases and data cubes.
The resulting analysis is a transparent script that can be followed to reproduce the
result from scratch. This allows learning from existing analyses and encourages readers
to engage with data.

@@

\begin{figure}[b]
\includegraphics[width=1\columnwidth]{figures/thegamma-walk}
\caption{Constructing a script that charts the top 5 members of the House of Lords for Kent, based
on their travel costs.}
\label{fig:walkthrough}
\end{figure}

@@

The review of related work suggests that there is an unexplored point in the design space of tools
for data exploration. Although various efforts make text-based programming easier, e.g. by providing
high-level declarative abstractions, most systems that target non-experts shy away from code, by using
either programming by example, natural language or graphical user interface. However, text-based
programming has notable benefits such as learnability and transparency. Our work is primarily
motivated by the question whether we can make text-based programming easy enough for non-experts?
The study presented in Section~\ref{sec:study} shows that this is, indeed, possible at least for typical data querying tasks.
The Gamma consists of a programming language, a web-based coding environment and a number of type providers
that enable access to various kinds of data sources. It allows non-experts to create programs using the
\emph{iterative prompting} interaction principle -- by repeatedly selecting an item from an auto-complete list.
%
We start with a walkthrough of the system (Section~\ref{sec:overview-walk}), before looking at details of
The Gamma language (Section~\ref{sec:overview-lang}) and type providers (Section~\ref{sec:overview-tps}).

\subsection{Querying Travel Expenses}
\label{sec:overview-walk}

To introduce The Gamma, we walk through a simple scenario.
A journalist from Kent is exploring travel expense claims by members of the House of Lords published
by the UK government. The data is available as a CSV file\footnote{Full data set has been obtained from
\url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }:

@@

\noindent
After the analyst imports the file through a web interface,
the environment is initialised with code that refers to the imported data as \ikvd{expenses}
and she starts exploring the data using the type provider for tabular data (Section~\ref{sec:overview-tps}),
following the steps illustrated in Figure~\ref{fig:walkthrough}:

@@

The Gamma uses a type system to infer what members are available at a given point in a chain.
Each expression has a type with a list of members that, in turn, have their own types.
The types are not built-in, but are generated by type providers for individual data sources.
The programming environment for The Gamma is based on the Monaco editor\footnote{For
more information, see \url{https://microsoft.github.io/monaco-editor/}}. When the user types `.'
the editor triggers auto-completion and retrieves a list of available members based on the type
information. The programming environment evaluates scripts on-the-fly and shows a preview as
illustrated in Figure~\ref{fig:thegamma}.

@@

The Gamma can be extended to support any kind of data source by implementing a \emph{type provider},
which defines a domain specific language for exploring data of a particular kind.
Technically, a type provider generates object types with members (such as \ikvd{paging} or \ikvd{Kent})
that are accessed via iterative prompting. We describe type provider for exploring data cubes
(inspired by an example from Syme et al. \cite{inforich}), tabular data (based on theory developed
by Petricek \cite{dotdriven}), and a novel type provider for exploring graph databases.

@@

\noindent
The dimensions of the \ikvd{worldbank} cube are countries, years and indicators, whereas the dimensions
of \ikvd{expenditure} are government services, years and value type (adjusted, nominal,
per GDP).  As illustrated in Figure~\ref{fig:cubetp}, the provider allows users to select a data series
from the data cube. Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
restricts the cube to a two-dimensional plane; selecting \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}
then gives a series with years as keys and emission data as values. Similarly, we could first filter the
data by a year or an indicator. The same mechanism is used to select UK government spending on
defence in terms of GDP over all available years.
As there is no widely used standard format for data cubes, adding another data cube to The Gamma
currently requires implementing a new type provider.

@@

Our second type provider allows users to construct queries to explore data in tabular formats
such as CSV files. Unlike the data cube provider, the provider for tabular data does not just
allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
the example from Figure~\ref{fig:thegamma}:

@@

\noindent
The example works with a CSV file that records individual medals awarded in Olympic games.
The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
calculates total number of gold and silver medals for each team (country) before sorting the data.

When using the provider, the user specifies a sequence of operations. Members such as
\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
determine the operation type. Those are followed by operation parameters. For example, when grouping
data, we first select the key and then choose a number of aggregations to calculate over the group.
Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
calculating the sum, average or the number of distinct values. As illustrated in
Section~\ref{sec:cases}, this is sufficient to construct a wide range of practical queries.

\subsubsection*{Graph Database Type Provider}
Our third type provider allows users to explore data from graph databases, which store
nodes representing entities and relationships between them. In the context of data journalism,
graph databases have been used for example in The Panama Papers reporting \cite{panama}.
The following example explores a database of Doctor Who characters and episodes. It retrieves
all enemies of the Doctor that appear in the Day of the Moon episode:

\begin{thegamma}
drwho.Character.Doctor
  .'ENEMY OF'.'[any]'.'APPEARED IN'.'Day of the Moon'
\end{thegamma}

@@

\begin{figure}[b]
\centering
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-mp}
  \caption{Exploring Olympic medallists using tabular data type provider}
  \label{fig:cases-mp}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-dr}
  \caption{Exploring Dr Who graph database by composing type providers}
  \label{fig:cases-dr}
\end{subfigure}
\vspace{-0.5em}
\caption{Charts produced by two case studies of using The Gamma.}
\label{fig:cases}
\end{figure}

@@

The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data visualizations. We consider the expressive power in
this section and assess simplicity in the next one. We used The Gamma to analyse
the UK government expenditure, activities of a research institute, Olympic medals and
information about the Doctor Who series\footnote{The analyses are available (non-anonymously)
at \url{http://gallery.thegamma.net}, \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.
In this section, we consider two interesting larger examples\footnote{Available
(non-anonymously) at: \url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.

\subsubsection*{If Michael Phelps were a Country}
Michael Phelps has won so many medals that media compared the number to
countries.\footnote{e.g.~``If Michael Phelps Were A Country, Where Would His Gold Medal Tally Rank?''
 \url{https://www.npr.org/sections/thetorch/2016/08/14/489832779/}}
Media illustrated the story with a chart showing a country league table including Michael Phelps.
We reproduce the chart, shown in Figure~\ref{fig:cases-mp}, using the tabular data type provider:

\begin{thegamma}
\kvd{let} data = olympics.'group data'.'by Team'.'sum Gold'.then
    .'sort data'.'by Gold descending'.then
    .paging.skip(43).take(4).'get series'.'with key Team'.'and value Gold'

\kvd{let} phelps = olympics.'filter data'.'Athlete is'.'Michael Phelps'.then
  .'group data'.'by Athlete'.'sum Gold'.then
  .'get series'.'with key Athlete'.'and value Gold'

charts.bar(data.append(phelps)).setColors(["#aec7e8","#aec7e8","#1f77b4"])
\end{thegamma}

\noindent
The data analysis is done in two commands. The first counts gold medals by countries
and uses paging to fetch 4 countries with suitable number of medals. In the second, we abuse the
grouping operation to aggregate data for just a single group. The two data series are then assigned
to local variables (for readability) and passed to the \ikvd{chart.columns} function.
The case study illustrates when more advanced language features are necessary. The data exploration
itself has been completed via iterative prompting, but producing the final chart currently requires
some manual programming. In practice, this would likely be done with the help of an expert
or by copying code from an example.

\subsubsection*{The Most Frequent Doctor Who Villains}
In the second case study, use a graph database on the Dr Who series and list Dr Who villains
by the number of episodes in which they appear. This case study is interesting as it
combines the graph database provider for fetching the data with the tabular data provider for
summarization:

@@

\noindent
Line 1 use the graph database provider to find all paths linking the Doctor node with any
character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
This produces a table that can be analysed using the tabular data provider by choosing the
\ikvd{explore} member. For each character (the villain) we count the total number of
distinct episodes in the table. The result is shown in Figure~\ref{fig:cases-dr}.
The example performs a fairly sophisticated data analysis that involves a graph database query
followed by an SQL-like data aggregation. The code can be fully constructed using iterative
prompting (with the exception of the numbers in paging). Another important feature is that
instantaneous previews support the user during the construction, showing the intermediate result
at each step of the construction.

@@

Our goal was to develop an easy-to-learn tool that non-programmers can use
for producing transparent data analyses. To evaluate the extent to which The Gamma achieves this,
we conducted a user study. We recruited participants among the operations and business team of a
non-profit research organization and investigate three research questions.

\vspace{0.5em}\noindent\emph{RQ1: Can non-programmers use The Gamma?}\hspace{0.3em}
Our first hypothesis is that non-programmers will be able to use The Gamma to explore data. To test
this, we gave participants one of four simple data exploration tasks and assess whether they were
able to complete the task, as well as how much assistance, if any, they needed.

\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}\hspace{0.3em}
Our second hypothesis is that users familiar with the iterative prompting principle will be able to
use an unfamiliar data source. We designed two of the four tasks to test this. Participants were
shown a demo using one data source and then asked to complete a task using a different one.

\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}\hspace{0.3em}
Our third hypothesis is that users can learn through percolation, i.e.~by looking at the source
code of published analyses. In one of our tasks, participants were given a minimal explanation
of The Gamma (showing how to initiate the iterative prompting process) together with
an extensive code sample.

\subsection{Study Design}
We performed a between-subjects study to evaluate the first experience of using The Gamma.
We recruited 13 participants (5 male, 8 female) from a business team
of a research lab working in non-technical roles (project management,
partnerships, communications) including one former journalist. Only one participant (\#12)
had prior programming experience. We split participants into 4 groups. We gave participants a
brief overview of The Gamma (with content depending on the task) and then asked participants
to complete a task. The four tasks were:

@@

\noindent
We let participants work independently, but offered guidance if they got stuck.
Tasks \emph{expenditure} and \emph{lords} aim to answer the questions RQ1 and RQ2; the task
\emph{worldbank} aims to answer RQ1 and RQ3. In \emph{olympics}, we test RQ1 using a more complex
data source and we ask further questions to explore the understanding of the data source. Following the
experiment, we conducted a short semi-structured group interview and later sent participants a follow-up
questionnaire.

\begin{figure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tabular}{l l l c l}
      \toprule
        & {\small \textit{Task}}
        & {\small \textit{Kind}} & {\small \textit{Done}}
        & {\small \textit{Notes}} \\
      \midrule
      \small \#1 & \small expenditure & \small cube & \priority{50} & {\small Obtained one data series}\\
      \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored further data }\\
      \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data  }\\
      \small \#4 & \small expenditure & \small cube & \priority{75}& {\small Correct member hint }\\
      \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data }\\
      \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
      \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
      \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
      \small \#9 & \small lords & \small table & \priority{75} & {\small  Composition issues}\\
      \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
      \small \#11 & \small lords & \small table & \priority{75} & {\small Hint to avoid operations}\\
      \small \#12 & \small olympics & \small table & \priority{75}  & {\small Hint to avoid operations}\\
      \small \#13 & \small olympics & \small table & \priority{75}  & {\small Two operations hints}\\
      \bottomrule
    \end{tabular}
    \caption{Work completed by individual participants; \priorityc{100} = completed,
      \priorityc{75} = required some guidance, \priorityc{50} = partially completed}
    \label{tab:tasks}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \begin{tabular}{p{17.25em} c c}
      \toprule
        {\small \textit{Questionnaire item}} & {\small \textit{Avg}} & {\small \textit{Sdv}} \\
      \midrule
      \small Considering your experience and possible use of The Gamma in the context of data journalism:\\
      \small \quad I found the system easy to use. & \small 4.00 & \small 0.89\\
      \small \quad Journalists will be able to use it to analyse data. & \small 3.45 & \small 1.04\\
      \small \quad Readers will be able to examine the analyses. & \small 3.73 & \small 0.79\\
      \small Thinking about the usability of the system:\\
      \small \quad Resulting programs are easy to understand. & \small 3.73 & \small 0.79\\
      \small \quad Text editor for creating them was easy to use. & \small 3.73 & \small 0.90\\
      \small \quad Finding items in the auto-complete was easy. & \small 3.45 & \small 1.37\\
      \small I would know how to explore a new data source\\
      \small \quad If given a comprehensive video tutorial. & \small 4.00 & \small 1.26\\
      \small \quad If given a number of code samples. & \small 3.36 & \small 1.12\\
      \small \quad Without any further guidance. & \small 2.45 & \small 1.13 \vspace{0.1em}\\
      \bottomrule
    \end{tabular}
    \caption{Follow-up questionnaire responses using
      a 5-point Likert scale (1 = strongly disagree, 5 = strongly agree) over 11 subjects.}
    \label{tab:quest}
  \end{subfigure}
  \vspace{-0.5em}
  \caption{Summary of the work completed by individual participants and responses to the follow-up questionnaire.}
  \label{tab:tabs}
  \vspace{-0.25em}
\end{figure}

\subsection{Study Results}
Figure~\ref{tab:tabs} shows the results of the study. Most notably, all participants were able to
complete, at least partially, a non-trivial data exploration task and only half of them required
further guidance. Table~\ref{tab:tasks} summarizes the work done by the study participants.
For each participant, we record the task, the kind of data source used in the task and the
level of completion. For participants who needed assistance, the notes section details the help
given. We discuss possible design improvements based on the experience in Section~\ref{sec:study-obs}.

The follow-up questionnaire was sent to all participants a week after the experiment. It was
completed all but two participants (\#8 and \#12). We asked general questions about the usability
of the system together with questions about learning designed to answer RQ2 and RQ3. A summary
of the answers is shown in Table~\ref{tab:quest}.

\subsection*{RQ1: Can non-programmers explore data with The Gamma?}
Three facts allow us to answer RQ1 in the affirmative. First, 9 out of 11 participants agree or
strongly agree that they ``found the system easy to use''. Second, participants spent 10--25 minutes
(average 17 minutes) working with The Gamma and 12 out of 13 completed the task; 6 required assistance,
but 3 of those faced one issue (discussed later) that could be addressed in the introduction.
Third, a number of participants shared positive comments in the interviews.
Participant \#3 found the system simple, but also points out an issue about data provenance,
which we revisit later:

\begin{quote}
\emph{``This is actually pretty simple to use. You think about the logic of what you're actually
  asking and then you try to get it into the format you can. But knowing where it comes from
  would tell you how to trust it.''}
\end{quote}

\noindent
Similarly, participant \#2, notes that The Gamma alleviated their unease about code:

\begin{quote}
\emph{``For somebody who does not do coding or programming, this does not feel that daunting.
  It's not like you're giving me large screen full of code, which is reassuring.''}
\end{quote}

\noindent
Finally, participant \#5 suggested the system could be used as an educational tool for teaching
critical thinking with data. They answer a follow-up question about what training materials would
the students need as follows:

\begin{quote}
\emph{``I don't think they'd need more than 5 minute video (..) this is the data
  source, this is what's in there.''}
\end{quote}

\subsection*{RQ2: Can knowledge transfer between data sources?}
Our study does not conclusively answer RQ2. There is some evidence in favor of a positive answer.
In the practical part, two of the tasks (\emph{expenditure} and \emph{lords}) used a different data
source in the introductory presentation than the one that the participants were asked to explore.
Participants were able to complete those tasks, although \emph{lords} has been more challenging
as it involves a more complex data source. In the interview, participant \#2 also gives a positive
answer:

\begin{quote}
\emph{``I found it quite easy to translate what you showed us in the demo to the new dataset.
   I thought it was quite easy to just get how that works.''}
\end{quote}

\noindent
Negative evidence is offered by the follow-up questionnaire. When asked
whether they would know how to approach a task using a new data source, 6 out of 11 participants
disagree or strongly disagree that they would know how to approach it without any further guidance.
The nature of RQ2 makes it a more challenging question to study and so finding a conclusive answer
arguably requires further research.

>>>>

\begin{abstract}
  Data exploration tools based on code have many desirable characteristics. They can easily access
  a wide range of different data sources, result in reproducible scripts and encourage users to
  reuse and modify existing data analyses. Unfortunately, most programming tools require expert
  coding skills. Can we make data exploration based on code accessible to non-experts?
  %
  We present The Gamma, a novel text-based data exploration environment.
  The Gamma is based on a single interaction principle and using it results in transparent and
  reproducible scripts. This lowers the barrier to entry and allows learning from previously
  created data analyses. We evaluate the usability and learnability of The Gamma through a user
  study on non-technical employees of a research institute.
  %
  Our work shows that we may not need to shy away from code in order to build accessible,
  reproducible and transparent tools that will allow a broad audience to benefit from the
  rise of open data.
\end{abstract}

@@

We contribute The Gamma, a text-based data exploration environment for non-experts. The Gamma
is based on a single, easy to understand interaction principle and provides a uniform
access to a range of data sources including data tables, graph databases and data cubes.
The resulting analysis is a transparent script that can be followed to reproduce the
result from scratch. This allows learning from existing analyses and encourages readers
to engage with data.

@@

\begin{figure}[b]
\includegraphics[width=0.94\columnwidth]{figures/thegamma-walk}
\caption{Constructing a script that charts the top 5 members of the House of Lords for Kent, based
on their travel costs.}
\label{fig:walkthrough}
\end{figure}

@@

The Gamma aims to make text-based data exploration easy enough for non-experts. The aim is
motivated, in part, by the desirable properties of text-based data exploration tools such
as transparency, reproducibility and learnability and, in part, by an aim to explore an
unexplored point in the design space of data exploration tools. Although various efforts
make text-based programming easier, most systems that target non-experts shy away from code.

The Gamma is a text-based data exploration environment that allows non-experts explore data using
iterative prompting -- by repeatedly selecting an item from an auto-complete list. The study
presented in Section~\ref{sec:study} confirms that the kind of data exploration shown in the next
section can, indeed, be successfully done by non-experts.
%
\subsection{Querying Travel Expenses}
\label{sec:overview-walk}

The walkthough in Figure~\ref{fig:walkthrough} shows a typical task completed using The Gamma.
A data analyst from Kent is exploring travel expense claims by members of the House of Lords published
by the UK government \cite{lords}. The following shows a subset of the data in the CSV format:

@@

\noindent
The analyst imports the file through a web interface, the environment is initialised with code
that refers to the imported data as \ikvd{expenses} and she starts exploring the data using the type provider for tabular data (Section~\ref{sec:overview-tps}):

@@

The Gamma uses a type system to infer what members are available at a given point in a chain.
Each expression has a type with a list of members that, in turn, have their own types.
The types are not built-in, but are generated by type providers for individual data sources.
The programming environment for The Gamma is based on the Monaco editor \cite{monaco}. When the user types `.'
the editor triggers auto-completion and retrieves a list of available members based on the type
information. The Gamma evaluates scripts on-the-fly and shows a preview as
illustrated in Figure~\ref{fig:thegamma}.

@@

The Gamma can be extended to support any data source by implementing a \emph{type provider},
which defines a domain specific language for exploring data of a particular kind.
A type provider generates object types with members (such as \ikvd{paging} or \ikvd{Kent})
that are accessed via iterative prompting. We describe type provider for exploring data cubes
(inspired by Syme et al. \cite{inforich}), tabular data (based on theory developed
by Petricek \cite{dotdriven}), and graph databases.

@@

\noindent
The dimensions of the \ikvd{worldbank} cube are countries, years and indicators, whereas the dimensions
of \ikvd{expenditure} are government services, years and value type (adjusted, nominal,
per GDP).  Figure~\ref{fig:cubetp} how the provider allows users to slice the data cube.
Choosing \ikvd{byCountry.\textquotesingle United States\textquotesingle},
restricts the cube to a plane and selecting \ikvd{\textquotesingle CO2 emissions (kt)\textquotesingle}
then gives a series with years as keys and emission data as values. Similarly, we could first filter the
data by a year or an indicator. The same mechanism is used to select UK government spending on
defence in terms of GDP.
%  As there is no widely used standard format for data cubes, adding another data cube to The Gamma
% currently requires implementing a new type provider.

@@

Our second type provider allows users to construct queries to explore data in tabular formats.
Unlike the data cube provider, the provider for tabular data does not just
allow selecting a subset of the data, but it can be used to construct SQL-like query. Consider
the example from Figure~\ref{fig:thegamma}:

@@

\noindent
The example queries a table that records individual medals awarded in Olympic games.
The chain constructs a query that selects rows corresponding to the Rio 2016 Olympics and then
calculates total number of gold and silver medals for each team (country) before sorting the data.
%
When using the provider, the user specifies a sequence of operations. Members such as
\ikvd{\textquotesingle filter data\textquotesingle} or \ikvd{\textquotesingle group data\textquotesingle}
determine the operation type. Those are followed by operation parameters. For example, when grouping
data, we first select the key and then choose a number of aggregations to calculate over the group.
Unlike SQL, the provider only allows users to choose from pre-defined aggregations such as
calculating the sum, average or the number of distinct values.
Section~\ref{sec:cases} shows that this is sufficient to construct a range of practical queries.

\subsubsection*{Graph Database Type Provider}
Our third type provider allows users to explore data from graph databases, which store
nodes representing entities and relationships between them.
The following example explores a database of Doctor Who characters and episodes. It retrieves
all enemies of the Doctor that appear in the Day of the Moon episode:

\begin{thegamma}
drwho.Character.Doctor.'ENEMY OF'.'[any]'.'APPEARED IN'.'Day of the Moon'
\end{thegamma}

@@

\begin{figure}[t]
\centering
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-dr}
  \caption{Exploring Dr Who graph database by composing type providers}
  \label{fig:cases-dr}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.49\textwidth}
  \centering
  \includegraphics[width=1\columnwidth]{figures/cases-mp}
  \caption{Exploring Olympic medallists using tabular data type provider}
  \label{fig:cases-mp}
\end{subfigure}
\vspace{-0.5em}
\caption{Charts produced by two case studies of using The Gamma.}
\label{fig:cases}
\end{figure}

@@

The Gamma aims to simplify programmatic data exploration while keeping enough expressive power
to allow users to create interesting data explorations. We assess the simplicity of The Gamma
in Section~\ref{sec:study}. Here, we consider two case studies that evaluate expressivity and show what
can be achieved using the simple iterative prompting principle\footnote{Available (non-anonymously) at:
\url{http://gallery.thegamma.net/86/} and  \url{http://gallery.thegamma.net/87/}, respectively.}.
We used The Gamma for larger projects that analyse the UK government expenditure, present
activities of a research institute, analyse Olympic medals and explore information about the Doctor
Who series\footnote{The analyses are available (non-anonymously)
at \url{http://gallery.thegamma.net}, \url{http://turing.thegamma.net} and \url{http://rio2016.thegamma.net}}.

\subsubsection*{The Most Frequent Doctor Who Villains}
Our first case study uses a graph database with data from the Dr Who series. It lists Dr Who villains
by the number of episodes in which they appear. This case study is interesting as it
combines the graph database provider for fetching the data with the tabular data provider for
summarization:

@@

\noindent
Line 1 use the graph provider to find all paths linking the Doctor with any
character linked via \ikvd{ENEMY OF}, followed by any episode linked by \ikvd{APPEARED IN}.
This produces a table that can be analysed using the tabular data provider by selecting
\ikvd{explore}. For each character (the villain) we count the number of
distinct episodes. The result is shown in Figure~\ref{fig:cases-dr}.
Despite performing a sophisticated data analysis that involves a graph database query,
followed by an SQL-like data aggregation, the code can be constructed using iterative
prompting, with the exception of the numbers in paging.

\subsubsection*{If Michael Phelps were a Country}
Michael Phelps has won so many medals that media compared the number to
countries~\cite{phelps}, often using a chart that shows a country league table including Michael Phelps
as an additional data point.
We reproduce the chart, shown in Figure~\ref{fig:cases-mp}, using the tabular data type provider:
%
\begin{thegamma}
\kvd{let} data = olympics.'group data'.'by Team'.'sum Gold'.then
    .'sort data'.'by Gold descending'.then
    .paging.skip(43).take(4).'get series'.'with key Team'.'and value Gold'
\vspace{5em}
\kvd{let} phelps = olympics.'filter data'.'Athlete is'.'Michael Phelps'.then
  .'group data'.'by Athlete'.'sum Gold'.then
  .'get series'.'with key Athlete'.'and value Gold'

charts.bar(data.append(phelps)).setColors(["#aec7e8","#aec7e8","#1f77b4"])
\end{thegamma}
%
The data analysis is done in three commands. The first counts gold medals by countries
and uses paging to fetch 4 countries with suitable number of medals. In the second, we use the
grouping operation to aggregate data for just a single group. The two data series are then assigned
to local variables (for readability) and passed to the \ikvd{chart.columns} function.
The example illustrates a case when more advanced language features are necessary. The data exploration
itself has been completed via iterative prompting, but producing the final chart currently requires
some manual programming. %In practice, this would likely be done with the help of an expert
%or by copying code from an example.

@@

Our objective is to develop a text-based data exploration tool that can be used by non-programmers.
Using the characterization by Olsen~\cite{evaluating}, data exploration environments are complex
systems that do not yield to simple controlled experimentation. We discuss the limits of evaluating
The Gamma in Section~\ref{sec:discussion-eval} and do not attempt to make make quantitative comparison
e.g.~with SQL or Tabelau~\cite{tableau}. Consequently, our goals are more modest.

\subsection{Study Design}
We evaluate the extent to which non-programmers can use The Gamma to conduct
text-based data exploration (\emph{RQ1}). To test this, we gave volunteers one of four data
exploration tasks and assessed whether they were able to complete the task, as well as how much
assistance, if any, they needed.
%
Some aspects of the study aim to shed light on three questions concerning learnability
of The Gamma, in particular whether knowledge can be trasferred between different data sources
(\emph{RQ2a}), whether users can learn from code samples without a detailed guidance
(\emph{RQ2b}) and to what extent they form a correct mental model of a more complex query
language used in the tabular data source (\emph{RQ2c}).

We performed a between-subjects study to evaluate the first experience of using The Gamma.
We recruited 13 participants (5 male, 8 female) from a business team
of a research lab working in non-technical roles (project management,
partnerships, communications) including one former journalist.  Only one participant (\#12)
had prior programming experience. We split participants into 4 groups. We gave participants a
brief overview of The Gamma (with content depending on the task) and then asked participants
to complete a task. The participants worked on the tasks for 30 minutes, after which we conducted
a 30 minute semi-structured group interview. We let participants work independently, but offered
guidance if they got stuck and asked for help. The four tasks were:
% The participants were volunteers who responded to an invitation sent to an internal mailing list.

@@

%\vspace{0.5em}\noindent\emph{RQ2: Can knowledge transfer between data sources?}\hspace{0.3em}
%Our second hypothesis is that users familiar with the iterative prompting principle will be able to
%use an unfamiliar data source. We designed two of the four tasks to test this. Participants were
%shown a demo using one data source and then asked to complete a task using a different one.
%
%\vspace{0.5em}\noindent\emph{RQ3: Can users learn from just code samples?}\hspace{0.3em}
%Our third hypothesis is that users can learn through percolation, i.e.~by looking at the source
%code of published analyses. In one of our tasks, participants were given a minimal explanation
%of The Gamma (showing how to initiate the iterative prompting process) together with
%an extensive code sample.
%

\noindent
All of the tasks aim to answer the primary question RQ1 for one of the supported data sources.
In addition, the tasks \emph{expenditure} and \emph{lords} study RQ2a because they use different
data source in introduction and taks; \emph{worldbank} study RQ2b in that it comes with only brief
introduction and \emph{olumpics} studies RQ2c using our most complex data source.

\begin{table}
\centering
\begin{tabular}{l l l c l}
  \toprule
    & {\small \textit{Task}}
    & {\small \textit{Kind}} & {\small \textit{Done}}
    & {\small \textit{Notes}} \\
  \midrule
  \small \#1\qquad\qquad & \small expenditure \qquad\qquad & \small cube \qquad\qquad & \qquad\priority{50}\qquad\qquad & {\small Obtained one of two data series}\\
  \small \#2 & \small expenditure & \small cube & \priority{100} & {\small Explored furhter data series }\\
  \small \#3 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
  \small \#4 & \small expenditure & \small cube & \priority{75}& {\small With hint to use another member }\\
  \small \#5 & \small expenditure & \small cube & \priority{100}& {\small Explored further data series }\\
  \small \#6 & \small worldbank & \small cube & \priority{75} & {\small With general syntax hint }\\
  \small \#7 & \small worldbank & \small cube & \priority{100} & {\small Completed very quickly }\\
  \small \#8 & \small worldbank & \small cube & \priority{100} & {\small Extra time to find data }\\
  \small \#9 & \small lords & \small table & \priority{75} & {\small  Struggled with composition}\\
  \small \#10 & \small lords & \small table & \priority{100} & {\small Completed very quickly }\\
  \small \#11 & \small lords & \small table & \priority{75} & {\small With a hint to avoid operations}\\
  \small \#12 & \small olympics & \small table & \priority{75}  & {\small With a hint to avoid operations}\\
  \small \#13 & \small olympics & \small table & \priority{75}  & {\small Hints about `then' and operations}\\
  \bottomrule
\end{tabular}

\caption{\centering Overview of work completed by individual participants in the study. \\The marks denote:
 \priorityc{100} = completed, \priorityc{75} = required some guidance, \priorityc{50} = partially completed}
\label{tab:tasks}
\vspace{-2em}
\end{table}

\subsection{Study Results}
Table~\ref{tab:tasks} summarizes the work done by the study participants. Most notably, all
participants were able to complete, at least partially, a non-trivial data exploration task and
only half of them required further guidance. For each participant, we record the task, the kind
of data source used in the task and the level of completion. For participants who needed assistance,
the notes section details the help given. The experience suggests a number of possible design
improvements for The Gamma, which are discussed below.

\subsubsection*{Can Non-programmers Explore Data with The Gamma?}
Three facts allow us to answer RQ1 in the affirmative. First, 9 out of 11 participants agree or
strongly agree that they ``found the system easy to use''. Second, participants spent 10--25 minutes
(average 17 minutes) working with The Gamma and 12 out of 13 completed the task; 6 required assistance,
but 3 of those faced one issue (discussed later) that could be addressed in the introduction.
Third, a number of participants shared positive comments in the interviews.
For example, participant \#3 noted taht \emph{``this is actually pretty simple to use.
You think about the logic of what you're actually asking and then you try to get it into the format you can.''}
Participant \#2 noted that The Gamma alleviated their unease about code:
\emph{``For somebody who does not do coding or programming, this does not feel that daunting.
  It's not like you're giving me large screen full of code, which is reassuring.''}

% \noindent
% Finally, participant \#5 suggested the system could be used as an educational tool for teaching
% critical thinking with data. They answer a follow-up question about what training materials would
% the students need as follows:
% \begin{quote}
% \emph{``I don't think they'd need more than 5 minute video (..) this is the data
%   source, this is what's in there.''}
% \end{quote}

\subsubsection*{How Users Learn The Gamma?}
The nature of RQ2 makes it a challenging question to study and finding conclusive answers
may require further research. There is some evidence that knowledge can be transferred between
different data sources (RQ2a). Two of the tasks (\emph{expenditure} and \emph{lords}) used
one data source in the introduction and another for the task. Participants were able to complete
those, although \emph{lords} has been more challenging as it involves a complex data source.
Participant \#2 also gives a positive answer \emph{``I found it quite easy to translate what you
showed us in the demo to the new dataset.''}.

There is also some evidence that users can learn just from code samples (RQ2b). In the
\emph{worldbank} task, participants were given only a minimal demo of how to invoke iterative
prompting together with print-out of 2 code samples. All three participants were able to complete
a related task using the same data source. When discussing suitable educational materials for
The Gamma, participant \#7 also confirmed that having code is sufficient when they noted that
\emph{``a video would just be this [i.e.~a code sample] anyway''}. This supports our hypothesis
that, once a user understands the iterative prompting principle, they can learn how to use any
specific data source just from code samples.

\subsubsection*{How Users Understand Complex Query Languages?}
Although iterative prompting itself is simple, specific data sources may
expose richer structure. To ascertain their understanding of the tabular type provider (RQ2c), we
asked participants who worked with tabular data (\emph{lords} or \emph{olympics}) about their
understanding of the  \ikvd{then} member. This is a regular member, but it has a
special meaning in the domain specific language. For example, when specifying a list of
aggregations, the user can repeatedly select further aggregations. Selecting \ikvd{then} completes
the operation and allows the user to choose other operations such as sorting.

Two participants (\#12 and \#13) initially thought that \ikvd{then} is used to split a command
over multiple lines, but rejected the idea after experimenting.
Participant \#12 then correctly concluded that it \emph{``allows us
to chain together the operations''} of the query; after a hint, participant \#13 reflected
that \emph{``if I knew this from the start, it would [have been easier].''} This summarizes an
important fact. Iterative prompting allows the users to start exploring new data sources, but
the structures exposed by more complex data sources have their own further design principles that
the users need to understand.
